---
title: "Assignment 3: Instrumental Variable Estimation"
subtitle: "Replicating the IV strategy in Stokes (2015)"
author: "EDS 241 / ESM 244 (DUE: 2/17/2026)"
format:
  html
date: "February 10, 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

------------------------------------------------------------------------

### Assignment instructions:

Working with classmates to troubleshoot code and concepts is encouraged. If you collaborate, list collaborators at the top of your submission.

All written responses must be written independently (in your own words).

Keep your work readable: Use clear headings and label plot elements thoughtfully (where applicable).

Submit both your rendered output and the Quarto (`.qmd`) file.

**Assignment submission (YOUR NAME)**: Joaquin Sandoval

------------------------------------------------------------------------

### Introduction

In this assignment, you will replicate the instrumental variable analysis from Stokes (2015), which examined how local wind turbine projects influenced electoral outcomes. Building on the matched dataset from Assignment 2, we will use Two-Stage Least Squares (2SLS) to estimate the causal effect of having a wind turbine proposed nearby on the change in Liberal Party vote share between 2007 and 2011. The instrument used in Stokes (2015) is a measure of local wind resource (average wind power, logged), which predicts where wind turbines are proposed. By using this instrument, we aim to isolate the portion of variation in turbine placement that is as-good-as-random, helping to meet the assumptions for causal identification.

**Study:** [Stokes, 2015 – Article](https://drive.google.com/file/d/1y2Okzjq2EA43AW5JzCvFS8ecLpeP6NKh/view?usp=drive_link)

**Data source:** [Dataverse – Stokes, 2015 replication data](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/SDUGCC)

------------------------------------------------------------------------

::: callout
Note: The estimates you obtain may not exactly match the published results in Stokes (2015) due to the alternative matching procedure used for processing the data in the previous assignment. Estimates should approximate the findings reported in *Table 2* of the article.
:::

------------------------------------------------------------------------

### Load packages

```{r}
library(tidyverse)
library(janitor)
library(here)
library(jtools)   # for export_summs (pretty regression tables)
library(AER)      # for ivreg (2SLS estimation)

```

------------------------------------------------------------------------

### Load the matched dataset (from Assignment 2)

> The matched_data has been preprocessed by matching on key covariates (e.g. pretreatment home values, education, income, population density) to improve balance between treated and control precincts. We will now use this data for the IV analysis. Make sure to re-code the `precinct_id` variable as a `factor`.

```{r}

matched_data <- read_csv(here::here("data" ,"matched_data_subset.csv")) |> 
    mutate(district_id = factor(district_id))

class(matched_data$district_id)

```

------------------------------------------------------------------------

### Part 1: IV Identification Rationale

Intuition for Using an Instrument:

**Question 1:** After matching on observables, why might we still need to utilize an instrumental variable approach to identify the causal effect of turbine proposals on vote share? In other words, what potential issues remain that an IV method can help address in this context? Use specific examples from the study to illustrate threats to a causal interpretation, then explain how an IV approach is designed to mitigate those threats.

*Response:* \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

After matching on observables, unobserved factors may drive both turbine placement and voting. An instrumental variable like wind speed isolates exogenous variation, identifying effects only where proposals occur due to wind resources, breaking the correlation between treatment and confounders

------------------------------------------------------------------------

### Part 2: Two-Stage Least Squares (2SLS) Step-Wise Implementation

#### 2A. First-Stage Estimation: Regress the treatment ($D$) on the instrument ($Z$)

$$D_i = \alpha_0 + \alpha_1 Z_i $$

a.  Estimate the first-stage regression of the treatment on the instrument (with controls). Regress `proposed_turbine_3km` on `log_wind_power`.
b.  Include the control variables used in Stokes (2015) for both stages: Distance to lakes, geographic coordinates (latitude & longitude) with their squares and interaction, plus district fixed effects.
c.  After running the first stage, report the F-statistic for the instrument.

```{r}
# Regress proposed_turbine_3km  on log_wind_power
first_stage <- lm(proposed_turbine_3km ~ log_wind_power + 
                      # Control variables
                  mindistlake + longitude + latitude + 
                      # Squares 
                  long_sq + lat_sq + 
                      # Interaction 
                      long_lat +
                      # Fixed effects of district_id  
                  district_id, 
                  data = matched_data)

export_summs(first_stage, digits = 3,
             model.names = c("First stage: Prpoposed Turbine 3km"),
             coefs = c("(Intercept)", "log_wind_power")) 

```

### Testing Instrument Relevance

Check instrument strength (F-statistic)

```{r}
first_stage_summary <- summary(first_stage)

first_stage_summary$fstatistic[1]

```

**Question 2A:** Based on the instrument relevance test reported in the study, would you conclude the instrument is strong enough to be credible? Explain what a weak instrument would mean in this setting: Specifically, what would it suggest about compliance with Ontario’s Green Energy Act policy?

*Response:* \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

The instrument is credible based on the relevance test. The study's F-statistic of 68 indicates wind power strongly predicts turbine placement. A weak instrument would suggest wind power does not strongly predict placement, indicating developers base proposals on other factors (local opposition, politics, land prices) rather than wind resources, which would imply low compliance with Ontario's Green Energy Act incentives.

------------------------------------------------------------------------

#### 2B. Second Stage Estimation

#### Regress the outcome ($Y$) on the fitted values from the 1st stage ($\hat{X}_i$)

$$Y_i = \beta_0 + \beta_1 \hat{D}_i + \epsilon_i$$

a.  Now estimate the second stage of the 2SLS.
b.  First, use the first-stage model to generate the predicted values of proposed_turbine_3km for each precinct (these are $\hat{D}_i$).
c.  Add these predicted values as a new column in matched_data (e.g. proposed_turbine_3km_HAT).
d.  Then, regress the outcome change_liberal (the change in Liberal vote share from 2007 to 2011) on the predicted treatment (proposed_turbine_3km_HAT), including the same controls and fixed effects as in the first stage.
e.  Fill in the code for these steps below to obtain the second-stage regression results.

Save predicted values $\hat{X}_i$ from first stage

```{r}
# Generate the predicted values of propsosed_turbin_3km for each precicnt 
matched_data$proposed_turbine_3km_HAT <- predict(first_stage)

```

#### Estimate the second-stage regression

$$LiberalVoteShare_i = \beta_0 + \beta_1 \widehat{ProposedTurbine}_i + ControlVariables... +  \epsilon_i$$

```{r}
second_stage <- lm(change_liberal ~ proposed_turbine_3km_HAT + 
                      # Control variables
                  mindistlake + longitude + latitude + 
                      # Squares and interaction 
                  long_sq + lat_sq + long_lat +
                      # Fixed effects of district_id  
                  factor(district_id), 
                  data = matched_data)

export_summs(second_stage, digits = 3, 
             model.names = c("Second stage: Change in Liberal Vote Share"),
             coefs = c("(Intercept)", "proposed_turbine_3km_HAT") ) 
```

#### Interpreting the 2SLS Estimate

**Question 2B:** Imagine you are explaining your 2SLS findings to a policymaker in Ontario. What does the estimated coefficient on `proposed_turbine_3km_HAT` imply about the electoral impact of a local wind turbine proposal (within 3 km) on liberal vote share?

**Question 2C:** Explain what it means that IV identifies a LATE in the context of the wind-turbine voting study. What specific subset of observations does the second-stage 2SLS estimate apply to, and what does it imply about interpretation and generalizability?

*Response:* \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

2B: Turbine proposals create electoral backlash against the Liberal Party, which implemented the Green Energy Act. Liberal vote share dropped by about 10 percentage points in precincts that received a turbine proposal, indicating significant NIMBY opposition in affected communities. 

2C: The IV identifies a Local Average Treatment Effect (LATE), meaning the causal effect applies only to precincts where turbine proposals were driven by wind availability, not other factors. This estimate applies to "wind-driven" locations responding to the policy. Regarding generalizability, the effect may differ in areas where placement is strategically targeted. For example, if developers avoid areas with expected opposition, the LATE may underestimate voting backlash.

------------------------------------------------------------------------

### Part 3: IV Assumptions and Validity

#### Evaluate Instrument Validity

**Question 3:** List the four key assumptions required for the IV strategy (2SLS) to identify a causal effect, and briefly explain what each one means in the context of this study. (Hint: think about what conditions a valid instrument must satisfy (relevance, exclusion,...)

*Response:* \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

1)  Relevance assumption: The instrument (wind power) must strongly predict the treatment (turbine proposals).

2)  Exclusion restriction: The instrument (wind power) affects the outcome (Liberal vote share) only through the treatment, not directly. Wind power must influence voting only by affecting proposals/turbine placement.

3)  Independence (Exogeneity) assumption: The instrument (wind power) is uncorrelated with unobserved confounders (e.g. baseline political preferences, local opposition).

4)  Monotonicity assumption: The instrument (wind power) affects treatment (proposals) in the same direction for all units. Higher wind power increases proposal likelihood for all precincts.

------------------------------------------------------------------------

### Part 4: Estimate 2SLS using `AER::ivreg()`

📜 [SEE Documentation for specification details: AER package Viggnette Example](https://cran.r-project.org/web/packages/AER/vignettes/AER.pdf)

::: callout-tip
Syntax for specifying 2SLS using `ivreg()`:

`ivreg( Y ~ D + CONTROLS  | Z + CONTROLS , data )`

-   The first-stage predictor variables go after the `~` symbol
-   The second-stage predictor variables go after the `|` symbol
:::

```{r}

# Before | : outcome ~ treatment + controls

fit_2sls <- ivreg(change_liberal ~ proposed_turbine_3km + 
                  mindistlake + longitude + latitude + 
                  long_sq + lat_sq + long_lat + factor(district_id) |
# After |: instrument + same controls
                  log_wind_power + 
                  mindistlake + longitude + latitude + 
                  long_sq + lat_sq + long_lat + factor(district_id),
                  data = matched_data)

export_summs(fit_2sls, digits = 3,
             model.names = c("Change in Liberal Vote Share"),
             coefs = c("(Intercept)", "proposed_turbine_3km")
             ) 
```

#### Robustness checking strategies utilized in Stokes, 2015

**Question 4:** Choose two *robustness checks* from the paper that the authors use to increase confidence in their causal identification strategy. For each one, summarize the logic and findings from the robustness check in your own words:

*Response:* \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

Distance Gradient Analysis:
If turbines cause electoral backlash, effects should be strongest near turbines and decline with distance. Proposed turbine effects were negative across all distances (0-5km) but not statistically distinguishable from each other. Operational turbine effects were large and significant from 0-3km, becoming indistinguishable from zero at 4-5km. This pattern supports causality rather than coincidence.

Voter Turnout Analysis:
Vote share decline could result from vote switching (Liberals defecting) or mobilization (anti-turbine voters showing out). Testing turnout isolates whether opposition created new participation. Turnout increased by 2.4 percentage points in proposed turbine precincts and 1.6 percentage points within 3km of proposals. Both mechanisms occurred: communities mobilized new voters and existing Liberal supporters defected.

------------------------------------------------------------------------

### END

------------------------------------------------------------------------
